### 2014-03-14 text/Chapter05.tex
Editing introduction chapter

\chapter{Coeffect meta-language} 
\label{ch:metalanguage}

%---------------------------------------------------------------------------------------------------

Introduction

@@

\section{Introduction}
\label{sec:metalanguage-intro}

?
\ExecuteMetaData[rules/rules.tex]{cml-types}

\section{Summary}
Oops!

>>>>

\chapter{Unified coeffect language} 
\label{ch:unified-coeffects}

% ==================================================================================================
% Macros - for ICFP 2014
% ==================================================================================================

% comonadic semantics
\newcommand{\F}{\mathsf{F}}
\newcommand{\D}{\mathsf{D}}
\newcommand{\Ds}[3]{\D^{\sclrd{#1}}_{#2} #3}
\newcommand{\Icat}{\cclrd{\mathbb{I}}}
\newcommand{\id}{\mathit{id}}
\newcommand{\B}[1]{\bar{#1}}
\newcommand{\bbullet}{\mathop{\bar{\bullet}}}
\renewcommand{\idf}[0]{ \ident{id} }
\newcommand{\m}{\mathsf{m}}
\newcommand{\n}{\mathsf{n}}

\newcommand{\shortar}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaro}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\shortarn}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaron}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\diagar}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}^{#4}"#1";"#2" }
\newcommand{\diagaro}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}_{#4}"#1";"#2" }

% Typing rule and typing statement (single line)
\newcommand{\tyrule}[3]{ \inference[(\footnotesize{#1})~]{#2}{#3} }
\newcommand{\tyruleCompact}[3]{ \text{(\footnotesize{#1})}~\dfrac{#2}{#3} }
\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
% Context transformation and context transformation with new line in the middle
\renewcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}
% Ordinary context where colouring is done by hand 
\newcommand{\coctx}[2]{ #1 \text{\scriptsize @} {\footnotesize{\text{$#2$}}}  }

% Simple coeffect systems in the introduction
\newcommand{\cons}{\hspace{-0.15em}\times\hspace{-0.25em}} %\sctxclrd{+\hspace{-0.55em}+}}

% Coeffect algebra
\definecolor{cclr}{rgb}{0.0,0.4,0.0}
\newcommand{\cclrd}[1]{ \textcolor{cclr}{#1} }

\newcommand{\C}{ {\textcolor{cclr}{\mathcal{C}}} }
\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} % \bullet} } % \textcolor{cclr}{\bullet}}  
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }

% \newcommand{\single}[{\langle{#1}\rangle}

\newcommand{\uncurry}{\Lambda^{\hspace{-0.2em}-1}}

% Coeffect shapes 
\definecolor{sclr}{rgb}{0.5,0,0.0}
\newcommand{\sclrd}[1]{ \textcolor{sclr}{#1} }

\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\renewcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }

\newcommand{\sflat}{\star}

% Coeffect algebra
\definecolor{aclr}{rgb}{0.0,0.0,0.7}
\newcommand{\aclrd}[1]{ \textcolor{aclr}{#1} }

\newcommand{\mksh}[1]{{\sclrd{[}#1\sclrd{]}}}

\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
% Tomas: I prefer \circledast here because it is a circled (lifted) version of \cseq 
% { \textcolor{aclr}{\bullet}}
\newcommand{\alift}[1]{ \textcolor{aclr}{\langle} #1 \textcolor{aclr}{\rangle}} 
\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }

% Free monoid used in the *structural* coeffect algebra
\newcommand{\fcons}{ \times } 
\newcommand{\fempty}{\epsilon}

@@

\section{Introduction}

\emph{Context} is important for defining meaning-- not just in natural languages, but also in logics and 
programming languages.
The standard notion of context in programming is an environment 
providing values for free variables. An open term with free variables is context dependent -- its 
meaning depends on the free-variable context. The simply-typed $\lambda$-calculus famously 
analyses such context usage. 
Other systems go further. %capturing how 
%the context is used. 
For example, bounded linear logic tracks the number of times a 
variable is used~\cite{girard1992bounded}.

In software engineering, ``context'' provides more than just free-variable values. For example, in a
distributed system, the context provides different resources that may be available on different 
devices (\eg{}, a database on a server or a GPS sensor on a phone).

In this paper we develop a calculus for capturing various notions of 
context in programming. A key feature and contribution
of the calculus is its \emph{coeffect system} which provides a static analysis for contextual properties (coeffects). The system
follows the style of type and effect systems, but captures a different class of properties. % (Section~\ref{sec:intro-pathways}). 
Another key contribution of the calculus is its semantics which can be smoothly instantiated for specific notions of context.
%The semantics of the calculus is general and can be instantiated for particular notions of context.

Coeffect systems were previously introduced as a generic analysis of
context dependence which can be instantiated for various notions of
context~\cite{petricek2013coeffects}. However, the formalization was restricted to
tracking a class of \emph{whole-context} properties where a term has just
one coeffect. This limited the applications and precision of any
analysis. For example, a whole-context liveness analysis marks the 
free-variable context as live (some variable may be used) or dead 
(no variable is used), but it cannot record liveness \emph{per-variable}.

We develop a more general system which captures both
\emph{per-variable} coeffects, which we call \emph{structural}, and
\emph{whole-context} coeffects, which we call \emph{flat}, and more. Our
key contributions are:

\begin{itemize}
\item We present a general coeffect type system (Section~\ref{sec:coeffects}) and demonstrate two 
  concrete uses -- \emph{flat} coeffect systems that tracks whole-context information and 
  \emph{structural} coeffect systems which tracks fine-grained per-variable information.

\item We show practical examples, instantiating the calculus for structural systems capturing 
  variable usage based on bounded linear logic, dataflow caching, and precise liveness analysis.
  We also instantiate the calculus to flat systems, building on and extending previous examples~\cite{petricek2013coeffects}.
% variants of the coeffect system capture examples described   in our earlier work

\item We discuss the syntactic properties of the \emph{structural} coeffect system 
  (Section~\ref{sec:syntax}). It satisfies type preservation under both $\beta$-reduction 
  and $\eta$-expansion, allowing its use with both call-by-name and call-by-value languages.
  This important property distinguishes it from both effect systems and flat coeffects.

\item We provide a denotational semantics, resisting and extending the notion of \emph{indexed comonads} to the 
structural setting (Section~\ref{sec:semantics}). We prove soundness by showing 
  the correspondence between syntactic and semantic properties of coeffect systems.
\end{itemize}

\noindent
Coeffects can be approached from multiple directions (Section~\ref{sec:intro-pathways}) including 
syntactic (effect systems), semantic, and proof-theoretic. We 
emphasize the syntactic view, though we also outline a categorical semantics
and note the interesting technical details.

% ==================================================================================================

\section{Why coeffects matter}
\label{sec:intro-coeffects}

Coeffects are way to describe notions of context in programming that keep turning up. 
To illustrate this, we overview three systems tracking contextual properties that 
 motivate our general coeffect system. Two systems track per-variable properties (bounded 
linear logic and dataflow) and one tracks whole-context properties (implicit parameters).
 We start with some background and finish with a brief overview of the literature leading to coeffects.

\subsection{Background, scalars and vectors}

The $\lambda$-calculus is asymmetric-- it maps a context with \emph{multiple} variables to a 
\emph{single} result. An expression with $n$ free variables of types $\sigma_i$ can be modelled
by a function
$\sigma_1 \times \ldots \times \sigma_n \rightarrow \tau$ with a product on the left, but a single value
on the right. Effect systems attach effect annotations to the result $\tau$. In coeffect systems,
we attach coeffects to the context $\sigma_1 \times \ldots \times \sigma_n$ and we often (but not always)
have one coeffect per each variable. We call the overall coeffect a \emph{vector} consisting of
\emph{scalar} coeffects. This asymmetry explains why coeffect systems are not trivially dual to 
effect systems.

It is useful to clarify how vectors are used in this paper. Suppose we have a set $\C$ of
\emph{scalars} such as $\cclrd{r_1},\ldots,\cclrd{r_n} \in \C$. A vector $\aclrd{R}$ over $\C$
is a tuple $\langle \cclrd{r_1}, \ldots, \cclrd{r_n}\rangle$ of scalars. 
We use letters like $\aclrd{R}, \aclrd{S}, \aclrd{T}$ for 
vectors and $\cclrd{r},\cclrd{s},\cclrd{t}$ for scalars.\footnote{For better readability, the paper 
distinguishes different structures using colours. However ignoring the colour does not introduce 
any ambiguity.} We also say that a \emph{shape} of a vector $\mksh{\aclrd{R}}$ (or more generally any container) 
is the set of \emph{positions} in a vector. So, a vector of length
$n$ has shape $\{ 1, 2, \ldots, n \}$.

Just as in scalar-vector multiplication, we lift any binary operation on scalars into a scalar-vector one:
$\cclrd{s} \bullet \aclrd{R} = \langle \cclrd{s}\bullet\cclrd{r_1}, \ldots, \cclrd{s}\bullet\cclrd{r_n}\rangle$.
Given two vectors $\aclrd{R},\aclrd{S}$ of the same shape, containing partially ordered scalars, we 
write $\aclrd{R} \leq \aclrd{S}$ for the pointwise extension of $\leq$ on scalars. Finally,
the associative operation $\times$ concatenates vectors.

We note that an environment $\Gamma$ containing $n$ uniquely named, typed variables is also a vector, 
but we continue to write `$,$' for the product, so $\Gamma_1, x\!:\!\tau, \Gamma_2$ should 
be seen as $\Gamma_1 \times \langle x\!:\!\tau\rangle \times \Gamma_2$.

\subsection{Bounded reuse}
\label{sec:intro-coeffects-bll}

Bounded linear logic \cite{girard1992bounded} restricts well-typed terms to polynomial-time algorithms.  
This is done by limiting the number of times a value (proposition) can be used. An assumption
$!_k A$ means that a variable can be used at most $k$ times. 
We attach annotations to the whole context rather than individual assumptions and so a context
$!_{k_1} A_1, ..., !_{k_n} A_n$ is written as $\coctx{\tau_1, ..., \tau_n}{\langle \cclrd{k_1}, ..., \cclrd{k_n}\rangle}$. 
This difference is further explained in Section~\ref{sec:related-work}. %For uniformity
%with later systems, we also write assumptions $A$ as $\tau$ and annotations $K$ as $\cclrd{r}$.

Bounded linear logic includes explicit weakening and contraction rules
that affect the multiplicity. Following the original logical style (but with our notation), these are written as:
\[
\inference
  {\coctx{\Gamma}{\aclrd{R}} \vdash \tau}
  {\coctx{\Gamma,\sigma}{\aclrd{R} \cons \alift{0}} \vdash \tau} 
\quad
\inference
  {\coctx{\Gamma_1,\sigma,\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
  {\coctx{\Gamma_1,\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \tau}
\]
The context $\coctx{\Gamma}{\aclrd{R}}$ includes a \emph{coeffect annotation} $\aclrd{R}$ which is a vector
$\alift{\cclrd{r_1}, \ldots, \cclrd{r_n}}$ of the same length as $\Gamma$ (a side-condition omitted for brevity).
In weakening (left), unused propositions are annotated with $0$ (no uses), while in contraction (right), multiple 
occurrences of a proposition are joined by adding the number of uses.

\begin{figure}[t]
\vspace{-0.5em}
\begin{align*}
\begin{array}{c}
%
\hspace{-0.6em}
\begin{array}{c}
\tyrule{var}{}
  {\coctx{x \!:\! \tau}{\alift{1}} \vdash x : \tau} 
\quad 
\tyruleCompact{weak}
  {\coctx{\Gamma}{R} \vdash e : \tau}
  {\coctx{\Gamma,x \!:\! \sigma}{\aclrd{R} \cons \alift{0}} \vdash e : \tau} 
\\[1.8em]
\tyruleCompact{sub}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R'}} \vdash e : \tau}~(\text{\scriptsize $\aclrd{R} \leq \aclrd{R'}$})
\quad 
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\aclrd{R} \cons \alift{s}} \vdash e : \tau}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau} 
\\[1.2em]
\end{array} \\
%
% ------------------------ 
%
\begin{array}{c}
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau \quad 
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} \ast \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\\[1.3em]
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\sigma,z\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons \alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s}+\cclrd{t}} \cons \aclrd{Q}} \vdash \subst{e}{z,y}{x} : \tau}
\\[1.3em]
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\sigma',y\!:\!\sigma,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x\!:\!\sigma',\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{t},\cclrd{s}} \cons \aclrd{Q}} \vdash e : \tau}
\end{array} 
% ------------------------ 
%
% \hspace{-1em}
%\begin{array}{c}
%
% TP: I removed the 'let' rule again because we do not need it to explain the storage axiom here.
% Also, I think the BLL sample is already used to explain quite a lot of things, so I'd be happier
% to keep it simple.
%
% BTW: The \tyruleCompact macro creates a rule that *looks* differently..
% (I do not understand why but it makes the line somehow thicker...)
%
% \tyruleCompact{let}
%   {\coctxs{\Gamma_1, v : \sigma}{R \times \langle{t}\rangle} \vdash e_1 : \tau \quad 
%    \coctxs{\Gamma_2}{S} \vdash e_2 : \sigma}
%   {\coctxs{\Gamma_1, \Gamma_2}{R \times (t \ast S)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau} \\[1.8em]
\end{array}
\end{align*}
\vspace{-0.5em}
\caption{type and coeffect system for bounded reuse}
\label{fig:bounded-coeff}
\end{figure}

\paragraph{Bounded linear coeffects.}
The system in Figure~\ref{fig:bounded-coeff} extends the outlined idea into a simple calculus.
Variable access (\emph{var}) has a singleton context with a singleton coeffect vector
$\alift{1}$. Weakening (\emph{weak}) extends the free-variable context with an unused variable and 
the coeffect with an associated scalar $0$. Explicit contraction (\emph{contr}) and
exchange (\emph{exch}) rules manipulate variables in the context and
modify the annotations accordingly -- adding the number of uses in
contraction and switching vector elements in exchange.

For abstraction (\emph{abs}), we know the number of uses of the parameter variable $x$ 
and attach it to the function type $\sigma \xrightarrow{\cclrd{s}}
\tau$ as a \emph{latent} coeffect. The remaining variables in $\Gamma$ are 
annotated with the remaining coeffect vector $\aclrd{R}$, specifying \emph{immediate} coeffects.

Application (\emph{app}) describes call-by-name evaluation.  Applying
a function that uses its parameter $\cclrd{t}$-times to an argument
that uses variables in $\Gamma_2$ $\aclrd{S}$-times means that, in
total, the variables in $\Gamma_2$ will be used $(\cclrd{t} \ast \aclrd{S})$-times. 
Recall that $\cclrd{t} \ast \aclrd{S}$ is a scalar multiplication of a
vector. Meanwhile, the variables in $\Gamma_1$ are used just
$\aclrd{R}$-times when reducing the expression $e_1$ to a function
value.

Finally, the sub-coeffecting rule (\emph{sub}) safely overapproximates
the number of uses using the pointwise
$\leq$ relation. We can view any variable as being used a greater
number of times than it actually is.

\paragraph{Example.} To demonstrate, consider a term
$(\lambda v.x+v+v)~(x+y)$. According to the call-by-name intuition, the variable $x$ is used three 
times -- once directly inside the function and twice via the variable $v$ after 
substitution. Similarly, $y$ is used 
twice. Assuming a judgment for the function body, abstraction yields:
%
\begin{equation*}
\tyrule{abs}
 { \coctx{x\!:\!\mathbb{Z},v:\mathbb{Z}}{\alift{\cclrd{1},\cclrd{2}}} \vdash x+v+v : \mathbb{Z} }
 { \coctx{x\!:\!\mathbb{Z}}{\alift{\cclrd{1}}} \vdash (\lambda v.x+v+v) : \mathbb{Z} \xrightarrow{\cclrd{2}} \mathbb{Z} }
\end{equation*}
%
To avoid name clashes, we $\alpha$-rename $x$ to $x'$ and later join $x$ and $x'$ using contraction.
Assuming $(x'+y)$ is checked in a context that marks $x'$ and $y$ as used once, the application rule yields
a judgment that is simplified as follows:
\[
\hspace{-0.7em}\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}
          {\alift{\cclrd{1}} \cons (\cclrd{2} \ast \alift{\cclrd{1},\cclrd{1}}) } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
{\inference
  { \coctx{x\!:\!\mathbb{Z},x'\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{1}, \cclrd{2}, \cclrd{2}} } \vdash (\lambda v.x+v+v)~(x'+y) : \mathbb{Z} }
  { \coctx{x\!:\!\mathbb{Z},y\!:\!\mathbb{Z}}{\alift{\cclrd{3}, \cclrd{2}}} \vdash (\lambda v.x+v+v)~(x+y)  : \mathbb{Z}} }
\]
%
The first step performs scalar multiplication, producing the vector
$\alift{\cclrd{1},\cclrd{2},\cclrd{2}}$. In the second step, we use contraction to join variables 
$x$ and $x'$ from the function and argument terms respectively.

It is worth pointing out that reduction by substitution yields $x+(x+y)+(x+y)$ which has the same coeffect as
the original. We return to evaluation strategies in Section~\ref{sec:syntax}, and 
show that structural coeffect systems preserve types and coeffects under $\beta$-reduction.

% --------------------------------------------------------------------------------------------------

\subsection{Dataflow and data access}

Dataflow languages such as Lucid \cite{wadge1985lucid} describe computations over \emph{streams}.
An expression is re-evaluated when new inputs are available (push) or when more output
is demanded (pull).  In causal dataflow, programs can access past
values of a stream. We consider a language where $\kvd{prev}~e$
returns the previous value of $e$, where 
$\kvd{prev}~(\kvd{prev}~e)$ therefore returns the second past value.

An implementation of causal dataflow may cache past values of
variables as an optimisation. The question is, how many past values
should be cached?  This can be approximated by a coeffect system.

\paragraph{Dataflow coeffects.} The coeffect system for dataflow is
similar to the one for bounded reuse in that it tracks a vector
of numbers $\aclrd{R}$ as part of the context
$\coctx{\Gamma}{\aclrd{R}}$. Here, coeffects represent the maximal number of
past values (\emph{causality depth}) required for a variable.

Weakening, exchange, abstraction and sub-coeffecting are the same as in bounded linear
coeffects, but the remaining rules differ. In Figure~\ref{fig:dataflow-coeff},
accessed variables (\emph{var}) are annotated with
$0$ meaning that no past value is required (only the
current one). The (\emph{prev}) rule crates caching requirements --
it increments the number of required values for all variables
used in $e$ using scalar-vector addition.

Application and contraction have the same structure as before, 
but use different operators. If two variables 
are contracting, requiring $\cclrd{s}$ and $\cclrd{t}$ past
values, then overall we need at most
$\ident{max}(\cclrd{s}, \cclrd{t})$ past values (\emph{contr}). That is, 
two caches are combined with the maximum of the two requirements, 
which satisfy the smaller requirements.

In (\emph{app}), the function requires $\cclrd{t}$ past values of
its parameter.  This means $\cclrd{t}$ past values of
$e_2$ are needed which in turn requires $\aclrd{S}$ past values of its free
variables $\Gamma_2$. Thus, we need $\cclrd{t}+\aclrd{S}$ past values of $\Gamma_2$ to perform the
call (\eg{}, we need $1+\aclrd{S}$ values to
get $1$ past value of the input $\sigma$, $2+\aclrd{S}$
values to get $2$ past values of $\sigma$, \emph{etc.}).

\paragraph{Example.} As an example, consider a function $\lambda x.\kvd{prev}~(y+x)$ applied to an argument
$\kvd{prev}~(\kvd{prev}~y)$. The body of the function accesses the past value of two variables, one free
and one bound:
\[
\dfrac
  {\coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1, 1}} \vdash \kvd{prev}~(y+x) : \mathbb{Z} }
  {\coctx{y\!:\!\mathbb{Z}}{\alift{1}} \vdash \lambda x . \kvd{prev}~(y+x) : \mathbb{Z} \xrightarrow{\cclrd{1}} \mathbb{Z} }
\]
The expression always requires the previous value of $y$ and adds it to
a previous value of the parameter $x$. Evaluating the value of the
argument $\kvd{prev}~(\kvd{prev}~y)$ requires two past values of $y$
and so the overall requirement is $3$ past values:
\[
\inference
  { \coctx{y\!:\!\mathbb{Z}}{\alift{1} } \vdash \lambda x.~(\ldots) \quad  \coctx{x\!:\!\mathbb{Z}}{\alift{2}} \vdash (\kvd{prev}~(\kvd{prev}~x) : \mathbb{Z} }
{\inference
  { \coctx{y\!:\!\mathbb{Z}, x\!:\!\mathbb{Z}}{\alift{1,3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~x)) : \mathbb{Z} }
  { \coctx{y\!:\!\mathbb{Z}}{\alift{3}} \vdash (\lambda x.\kvd{prev}~(y+x))~(\kvd{prev}~(\kvd{prev}~y)) : \mathbb{Z} } }
\]
The derivation uses (\emph{app}) to get requirements $\alift{1,3}$ and then 
(\emph{contr}) to take the maximum, showing three past values are sufficient. Reducing 
the expression by substitution we get $\kvd{prev}~(y+(\kvd{prev}~(\kvd{prev}~y)))$. 
Semantically, this performs stream lookups $y[1] + y[3]$ where the indices are the 
number of enclosing $\kvd{prev}$s.% expressions.

We previously used dataflow as an example of
coeffects~\cite{petricek2013coeffects}, but tracked caching
requirements on the whole context. The system outlined here is more
powerful and practically useful, with finer-grained coeffects tracking
caching requirements per-variable.

% --------------------------------------------------------------------------------------------------

\begin{figure}[t]
\[
\tyrule{contr}
  {\coctx{\Gamma_1,y\!:\!\tau,z\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\cclrd{s},\cclrd{t}} \cons \aclrd{Q}} \vdash e : \tau}
  {\coctx{\Gamma_1,x\!:\!\tau,\Gamma_2}{\aclrd{R}\cons\alift{\ident{max}(\cclrd{s},\cclrd{t})} \cons \aclrd{Q}} \vdash \subst{e}{y,z}{x} : \tau}
\]
\[
\tyrule{app}
  { \coctx{\Gamma_1}{\aclrd{R}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
  { \coctx{\Gamma_1, \Gamma_2}{\aclrd{R} \cons (\cclrd{t} + \aclrd{S})} \vdash e_1 \, e_2 : \tau} 
\]
\[
\tyrule{var}{}
  {\coctx{x\!:\!\tau}{\alift{0}} \vdash x : \tau}
\quad\;
\tyrule{prev}
  {\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau}
  {\coctx{\Gamma}{1 + \aclrd{R}} \vdash \kvd{prev}~e : \tau}  
\]
\caption{type and coeffect system for dataflow caching}
\label{fig:dataflow-coeff}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Implicit parameters}
\label{sec:intro-coeffects-impl}

As our third example, we revisit Haskell implicit parameters \cite{lewis2000implicit} 
used in our earlier coeffect work \cite{petricek2013coeffects}. Implicit parameters are variables
that mix aspects of dynamic and lexical scoping. Implicit parameters are a distinct syntactic 
category to variables and we write them as \ident{?p}. For simplicity, we omit \emph{let}-binding for
implicit parameters and focus just on tracking requirements.

\paragraph{Implicit parameters coeffects.} Implicit parameters are a whole-context coeffect 
not linked to ordinary variables. We keep track of sets of implicit parameters that are required
by an expression (and their types). For example $\coctx{\Gamma}{\{ \ident{?p}_1 : \tau_1, \ldots, \ident{?p}_n : \tau_n \}}$
means that a context provides ordinary variables $\Gamma$ and values for implicit parameters $\ident{?p}_i$.
Unlike in the previous examples, we no longer need to distinguish between coeffects attached
to variables (scalars) and coeffects attached to contexts (vectors), so we write $\cclrd{r},\cclrd{s},\cclrd{t}$
for both.

Despite the differences, the type system in Figure~\ref{fig:implpar-coeff} follows the same 
structure as the earlier two examples. Context requirements are created when accessing an implicit
parameter (\emph{param}) (a system-specific rule). Structural rules (exchange, weaken, contract) 
do not affect the coeffects. For example parameters are reordered in (\emph{exch}),
but this has no effect as set union $\cup$ is commutative.

In abstraction and application, the structural $\cons$ operator (previously vector 
concatenation) becomes $\cup$. Sets of implicit parameters are not associated to
individual variables and so they are unioned. The (\emph{app}) rule uses $\cup$ to
combine the implicit parameters required by the function with the requirements of the argument too.

We call this a \emph{flat} coeffect system since coeffects have only
one shape (there is no scalar/vector distinction).  Other flat
coeffect systems may use a richer structure~\cite{petricek2013coeffects}. In
particular, the operations used in abstraction and application may differ
(to accommodate over-approximation). We return to this in
Section~\ref{sec:coeffects-structural}.

\paragraph{Example.} Unlike structural coeffect systems, flat systems
do not necessarily have principal coeffects. 
This arises from the (\emph{abs}) rule which can freely split
requirements between the function type and the declaring
context. Consider a function %filters news based on a location:
$\lambda ().\ident{?p}_1 + \ident{?p}_2$.  There are nine possible 
 type and coeffect derivations, two of which are:
%% 1 p1p2  p1p2
%% 2 p1p2  p1
%% 3 p1p2  p2
%% 4 p1p2  
%% 5 p1    p1p2
%% 6 p1    p2
%% 7 p2    p1p2
%% 8 p2    p1
%% 9       p1p2
\[
\begin{array}{rl}
\coctx{\emptyset}{ \{  \} } &\hspace{-0.7em}\vdash (\ldots) : 
  \ident{unit} \xrightarrow{ \{ \ident{?p}_1:\mathbb{Z}, \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
\\
\coctx{\emptyset}{ \{ \ident{?p}_1:\mathbb{Z} \} } &\hspace{-0.7em} \vdash (\ldots) : 
  \ident{unit} \xrightarrow{ \{ \ident{?p}_2:\mathbb{Z} \} } \mathbb{Z}
\end{array}
\]
In the first case, both parameters are dynamically scoped and have to be provided by the caller.
In the second case, the parameter $\ident{?p}_1$ is available in the declaring scope and so it is
(lexically) captured.

Although structural coeffects have more desirable syntactic properties, we aim to capture this
non-principality too as it is practically useful -- Haskell's implicit parameters use it and
it can be used to model resource rebinding in distributed systems such as~\cite{app-distributed-acute}).

% --------------------------------------------------------------------------------------------------

\begin{figure}
\[
\tyrule{exch}
  {\coctx{\Gamma_1,x\!:\!\tau,y:\sigma,\Gamma_2}{r \cup \cclrd{s} \cup \cclrd{t} \cup \cclrd{q}} \vdash e : \tau}
  {\coctx{\Gamma_1,y\!:\!\sigma,x:\tau,\Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s} \cup \cclrd{q}} \vdash e : \tau}
\]
\[
\tyrule{app}
  { \coctx{\Gamma_1}{\cclrd{r}} \vdash e_1 : \sigma \xrightarrow{\cclrd{t}} \tau &
    \coctx{\Gamma_2}{\cclrd{s}} \vdash e_2 : \sigma }
  { \coctx{\Gamma_1, \Gamma_2}{\cclrd{r} \cup \cclrd{t} \cup \cclrd{s}} \vdash e_1 \, e_2 : \tau} 
\]
\[
\tyrule{param}{}
  {\coctx{()}{ \{ \ident{?p}:\tau \} } \vdash \ident{?p} : \tau}
\quad\;
\tyrule{abs}
  {\coctx{\Gamma, x \!:\! \sigma}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x . e : \sigma \xrightarrow{\cclrd{s}} \tau}
\]
\vspace{-0.5em}
\caption{Type and coeffect system for implicit parameters}
\label{fig:implpar-coeff}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Pathways to coeffects}
\label{sec:intro-pathways}

This paper mainly follows work on effect systems and their link to categorical semantics. We briefly 
review this and other directions leading to coeffects. An eager reader can return to this section later.

\paragraph{Effect systems.}

Effect systems \cite{gifford1986effects} track
effectful operations of computations such as memory access or lock
usage \cite{flanagan2003type}. They are written as judgments $\Gamma
\vdash e : \tau ~\&~\cclrd{\rho}$ associating effects
$\cclrd{\rho}$ with the result. Effect systems capture \emph{output
  effects} where, as Tate puts it, \emph{``all computations with [an]
  effect can be thunked as pure computations for a domain-specific
  notion of purity.''}~\cite{effects-producer-semantics}.  This
thunking is typically a $\lambda$-abstraction. Given an effectful
expression $e$, the function $\lambda x . e$ is an effect-free value
that delays all effects:
%
\begin{equation*}
\inference[({abs})\;]
  {\Gamma, x\!:\!\tau_1 \vdash e : \tau_2 ~\&~\cclrd{\rho}} 
  {\Gamma \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{\rho}} \tau_2~\&~\cclrd{\emptyset}}
\end{equation*}
%
Coeffects do not follow this pattern.  In contrast to effect systems,
context requirements cannot be easily ``thunked'' as
pure values. Lambda abstraction can split context requirements
between \emph{immediate} and \emph{latent} requirements. This is akin
to how lambda abstraction splits a free-variable context into 
the bound parameter (call site) and the remaining free variables (declaration site).

\paragraph{Categorical semantics.}
Moggi models effectful computations as functions of type 
$\tau_1 \rightarrow \mtyp{}{\tau_2}$ where $M$ is monad providing composition of effectful 
computations~\cite{moggi1991monads}. Wadler and Thiemann \cite{wadler2003marriage} link effect systems with monads using 
annotated monads $\tau_1 \rightarrow \mtyp{\cclrd{\rho}}{\tau_2}$ whose semantics has been 
provided by Katsumata~\cite{katsumata2014parametric}.

Context-dependent computations require a different model. Uustalu and Vene \cite{uustalu2008notions} 
use functions $C{}{\tau_1} \rightarrow \tau_2$ where $C$ is a \emph{comonad}. Our earlier work 
\cite{petricek2013coeffects} used indexed comonads with denotations $C^{\cclrd{r}}{\tau_1} \rightarrow \tau_2$ adding 
annotations akin to Wadler and Thiemann. In Section~\ref{sec:semantics} we extend
indexed comonads to capture the general coeffect systems of this paper,  
in the style of Katsumata. 
% 
% TP: Mentioned Katsumata here. Thought it is nice to say we build on new work here (as opposed to ICALP)
%

\paragraph{Language and meta-language.}
Moggi uses monads in two systems~\cite{moggi1991monads}. In the first system, a monad is used to model
an effectful language itself -- the semantics of a language uses a specific monad. In the second system, 
monads are added as type constructors, together with syntax corresponding to \emph{unit} and \emph{bind}
operations.
